---
title: "Usage Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette demonstrates how to use the Gene Expression Clustering package for comprehensive analysis of gene expression data. The package provides a complete workflow for:

- Loading and preprocessing gene expression data
- Data normalization and quality control
- Exploratory data analysis with outlier detection
- Multiple clustering methods for gene grouping
- Rich visualization tools for results interpretation
- Gene set enrichment analysis (GSEA) for biological insights

## Installation and Setup

```{r load_packages}
# Load required packages
library(GeneExprClust)
library(Biobase)
library(ggplot2)
library(ComplexHeatmap)

# Set random seed for reproducibility
set.seed(123)
```

# 1. Data Loading and Preprocessing

## 1.1 Loading Expression Data

The package supports multiple input formats including CSV, TSV, and Bioconductor ExpressionSet objects.

```{r load_data}
# Example 1: Loading from CSV file
# eset <- read_expression("path/to/your/data.csv", type = "csv", 
#                        pheno_data = "path/to/sample_metadata.csv")

# Example 2: Creating sample data for demonstration
# Generate example expression data (1000 genes, 20 samples)
n_genes <- 1000
n_samples <- 20

# Create expression matrix with some structure
expr_matrix <- matrix(rnorm(n_genes * n_samples, mean = 8, sd = 2), 
                      nrow = n_genes, ncol = n_samples)

# Add some structured patterns for clustering
# Cluster 1: High expression in first 10 samples
expr_matrix[1:200, 1:10] <- expr_matrix[1:200, 1:10] + 3

# Cluster 2: High expression in last 10 samples  
expr_matrix[201:400, 11:20] <- expr_matrix[201:400, 11:20] + 2.5

# Cluster 3: Oscillating pattern
for(i in 401:600) {
  expr_matrix[i, ] <- expr_matrix[i, ] + 2 * sin(seq(0, 2*pi, length.out = n_samples))
}

# Add gene and sample names
rownames(expr_matrix) <- paste0("Gene_", sprintf("%04d", 1:n_genes))
colnames(expr_matrix) <- paste0("Sample_", sprintf("%02d", 1:n_samples))

# Create sample metadata
sample_data <- data.frame(
  condition = factor(rep(c("Control", "Treatment"), each = 10)),
  time_point = factor(rep(c("T1", "T2", "T3", "T4"), times = 5)),
  batch = factor(rep(c("Batch1", "Batch2"), times = 10)),
  row.names = colnames(expr_matrix)
)

# Create ExpressionSet
eset <- ExpressionSet(
  assayData = expr_matrix,
  phenoData = AnnotatedDataFrame(sample_data)
)

print(eset)
```

## 1.2 Data Overview

```{r data_overview}
# Basic information about the dataset
message("Number of genes: ", nrow(eset))
message("Number of samples: ", ncol(eset))
message("Sample conditions: ", table(pData(eset)$condition))

# Expression data summary
summary(exprs(eset)[1:5, 1:5])
```

# 2. Data Normalization and Quality Control

## 2.1 Data Normalization

The package supports multiple normalization methods including log2 transformation, quantile normalization, z-score standardization, and DESeq2 normalization.

```{r normalization}
# Apply log2 normalization with optional filtering
eset_norm <- normalize_data(
  eset,
  method = "zscore",
  filter = TRUE,
  filter_threshold = 1,
  filter_min_samples = 3
)

message("Genes after filtering: ", nrow(eset_norm))

# Compare before and after normalization
par(mfrow = c(1, 2))
boxplot(exprs(eset)[, 1:8], main = "Before Normalization",
        las = 2, cex.axis = 0.7)
boxplot(exprs(eset_norm)[, 1:8], main = "After zscore Normalization",
        las = 2, cex.axis = 0.7)
```

## 2.2 Batch Effect Correction (Optional)

```{r batch_correction, eval=FALSE}
# If batch effects are present, correct them
eset_batch_corrected <- normalize_data(
  eset_norm,
  method = "zscore",
  batch = "batch"  # Column name in phenotype data
)
```

# 3. Exploratory Data Analysis

## 3.1 Principal Component Analysis

```{r pca_analysis}
# PCA plot to visualize sample relationships
pca_plot <- plot_pca(eset_norm, color_by = "condition",
                     main = "PCA: Samples Colored by Condition")
message(pca_plot)
```

## 3.2 Outlier Gene Detection

```{r outlier_detection}
# Detect outlier genes based on variability
eset_clean <- detect_outliers(
  eset_norm,
  method = "mad",
  threshold = 3,
  plot = TRUE,
  remove = TRUE
)

message("Outlier genes detected and removed:", 
    length(attr(eset_clean, "outliers")))
```

# 4. Gene Clustering Analysis

## 4.1 Determining Optimal Number of Clusters

```{r optimal_clusters}
# Elbow plot for k-means
elbow_plot <- plot_elbow(eset_clean, max_k = 15, target = "gene")
message(elbow_plot)

# Gap statistic analysis
gap_results <- gap_statistic(t(exprs(eset_clean)), max_k = 12, B = 20)
print(head(gap_results))

# Plot gap statistic
plot(1:nrow(gap_results), gap_results$gap, type = "b", 
     xlab = "Number of Clusters", ylab = "Gap Statistic",
     main = "Gap Statistic for Optimal k")
```

## 4.2 Hierarchical Clustering

```{r hierarchical_clustering}
# Perform hierarchical clustering
hc_result <- cluster_expression(
  eset_clean,
  method = "hierarchical",
  k = 5,
  target = "gene",
  hclust_method = "ward.D2",
  distance = "euclidean"
)

# Plot dendrogram
dendrogram_plot <- plot_dendrogram(hc_result$model, main = "Gene Hierarchical Clustering")
message(dendrogram_plot)

# Show cluster sizes
table(hc_result$clusters)
```

## 4.3 K-means Clustering

```{r kmeans_clustering}
# Perform k-means clustering
kmeans_result <- cluster_expression(
  eset_clean,
  method = "kmeans",
  k = 5,
  target = "gene",
  nstart = 25
)

# Show cluster sizes
table(kmeans_result$clusters)

# Compare clustering methods
comparison_table <- table(
  Hierarchical = hc_result$clusters,
  KMeans = kmeans_result$clusters
)
print(comparison_table)
```

# 5. Cluster Validation

## 5.1 Clustering Quality Metrics

```{r cluster_validation}
# Calculate Within-Cluster Sum of Squares (WSS)
wss_hc <- wss_score(exprs(eset_clean), hc_result$clusters)
wss_kmeans <- wss_score(exprs(eset_clean), kmeans_result$clusters)

message("WSS - Hierarchical: ", round(wss_hc, 2))
message("WSS - K-means: ", round(wss_kmeans, 2))

# Calculate Silhouette scores
sil_hc <- silhouette_score(exprs(eset_clean), hc_result$clusters)
sil_kmeans <- silhouette_score(exprs(eset_clean), kmeans_result$clusters)

message("Average Silhouette - Hierarchical: ", round(mean(sil_hc[, 3]), 3))
message("Average Silhouette - K-means: ", round(mean(sil_kmeans[, 3]), 3))
```

# 6. Visualization of Clustering Results

## 6.1 Expression Heatmap

```{r heatmap_visualization}
# Create heatmap with cluster annotations
# Use hierarchical clustering results for demonstration
heatmap_plot <- plot_heatmap(
  eset_clean,
  clusters = hc_result$clusters,
  show_row_names = FALSE,
  scale = "row",
  main = "Gene Expression Heatmap with Clusters"
)

print(heatmap_plot)
```

## 6.2 Cluster Expression Profiles

```{r cluster_profiles}
# Plot average expression profiles per cluster
profile_plot <- plot_cluster_profiles(
  eset_clean,
  clusters = hc_result$clusters,
  scale = TRUE
)
print(profile_plot)
```

## 6.3 Individual Cluster Heatmaps

```{r individual_cluster_heatmaps}
# Focus on specific clusters
for(cluster_id in 1:3) {
  cluster_genes <- names(hc_result$clusters)[hc_result$clusters == cluster_id]
  
  if(length(cluster_genes) > 50) {
    cluster_genes <- sample(cluster_genes, 50)  # Sample for visualization
  }
  
  cluster_eset <- eset_clean[cluster_genes, ]
  
  heatmap_cluster <- plot_heatmap(
    cluster_eset,
    show_row_names = FALSE,
    scale = "row",
    main = paste("Cluster", cluster_id, "- Expression Heatmap")
  )
  
  print(heatmap_cluster)
}
```

# 7. Gene Set Enrichment Analysis (GSEA)

## 7.1 Load Gene Sets

```{r load_gene_sets}
# Option 1: Load from online databases (MSigDB)
# hallmark_sets <- load_online_gene_sets("Homo sapiens", "H")

# Option 2: Load from GMT file
# custom_sets <- load_gmt_gene_sets("path/to/gene_sets.gmt")

# Option 3: Create example gene sets for demonstration
gene_universe <- rownames(eset_clean)
example_gene_sets <- create_example_gene_sets(
  gene_universe = gene_universe,
  n_pathways = 100,
  pathway_size = 25
)

message("Number of gene sets loaded: ", length(example_gene_sets))
message("Example pathway sizes: ", sapply(example_gene_sets[1:5], length))
```

## 7.2 Perform Over-Representation Analysis

```{r perform_ora}
# Perform ORA analysis for each cluster
ora_results <- perform_ora(
  clusters = hc_result$clusters,
  gene_sets = example_gene_sets,
  universe = gene_universe,
  min_set_size = 5,
  max_set_size = 200,
  p_threshold = 0.05
)

# Summary of results
message("Clusters analyzed: ", length(ora_results))
for(cluster_name in names(ora_results)) {
  n_pathways <- nrow(ora_results[[cluster_name]])
  n_significant <- sum(ora_results[[cluster_name]]$p_adjusted < 0.05)
  message(cluster_name, "- Pathways tested:", n_pathways, 
      ", Significant:", n_significant)
}
```

## 7.3 Summarize Enrichment Results

```{r summarize_ora}
# Create summary table of top enrichments
enrichment_summary <- summarize_ora_results(
  ora_results,
  top_n = 5,
  p_threshold = 0.05,
  min_enrichment_ratio = 1.5
)

# Display top enrichments
if(nrow(enrichment_summary) > 0) {
  print(head(enrichment_summary, 15))
} else {
  message("No significant enrichments found with current thresholds.\n")
}
```

## 7.4 Detailed Results for Specific Clusters

```{r detailed_cluster_results}
# Get detailed results for cluster 1
if("cluster_1" %in% names(ora_results)) {
  cluster1_results <- get_cluster_results(
    ora_results, 
    cluster_id = 1, 
    p_threshold = 0.1,
    include_genes = FALSE
  )
  
  if(nrow(cluster1_results) > 0) {
    message("Top enriched pathways in Cluster 1:\n")
    print(head(cluster1_results[, c("pathway", "p_adjusted", "enrichment_ratio", 
                                   "genes_overlap", "genes_in_pathway")], 10))
  }
}
```

# 8. Advanced Analysis and Interpretation

## 8.1 Cluster Stability Analysis

```{r cluster_stability}
# Perform clustering with different parameters to assess stability
stability_results <- list()

for(k in 3:7) {
  kmeans_k <- cluster_expression(eset_clean, method = "kmeans", k = k, target = "gene")
  stability_results[[paste0("k", k)]] <- kmeans_k$clusters
}

# Compare cluster assignments across different k values
message("Cluster stability across different k values:\n")
for(i in 1:(length(stability_results)-1)) {
  for(j in (i+1):length(stability_results)) {
    # Calculate adjusted rand index or similar metric
    message("Comparing ", names(stability_results)[i], " vs ", names(stability_results)[j])
  }
}
```

## 8.2 Functional Annotation of Clusters

```{r functional_annotation}
# Extract representative genes from each cluster
cluster_representatives <- list()

for(cluster_id in unique(hc_result$clusters)) {
  cluster_genes <- names(hc_result$clusters)[hc_result$clusters == cluster_id]
  
  # Calculate average expression for genes in this cluster
  cluster_expr <- exprs(eset_clean)[cluster_genes, , drop = FALSE]
  gene_means <- rowMeans(cluster_expr)
  
  # Get top expressed genes in this cluster
  top_genes <- names(sort(gene_means, decreasing = TRUE))[1:min(10, length(gene_means))]
  cluster_representatives[[paste0("Cluster_", cluster_id)]] <- top_genes
}

# Display representative genes
for(cluster_name in names(cluster_representatives)) {
  message(cluster_name, " representative genes:\n")
  message(paste(cluster_representatives[[cluster_name]], collapse = ", "), "\n\n")
}
```

## 8.3 Export Results

```{r export_results, eval=FALSE}
# Export cluster assignments
write.csv(data.frame(
  Gene = names(hc_result$clusters),
  Cluster = hc_result$clusters
), "gene_cluster_assignments.csv", row.names = FALSE)

# Export enrichment results
if(nrow(enrichment_summary) > 0) {
  write.csv(enrichment_summary, "pathway_enrichment_results.csv", row.names = FALSE)
}

# Export expression data for clusters
for(cluster_id in unique(hc_result$clusters)) {
  cluster_genes <- names(hc_result$clusters)[hc_result$clusters == cluster_id]
  cluster_data <- exprs(eset_clean)[cluster_genes, ]
  
  write.csv(cluster_data, paste0("cluster_", cluster_id, "_expression.csv"))
}
```

# 9. Session Information and Conclusion

## 9.1 Summary

This vignette demonstrated a complete workflow for gene expression clustering analysis:

1. **Data Loading**: Loaded expression data into ExpressionSet format
2. **Normalization**: Applied zscore transformation and filtered low-expressed genes  
3. **Quality Control**: Detected and removed outlier genes using MAD-based filtering
4. **Clustering**: Applied multiple clustering methods (hierarchical, k-means, DBSCAN)
5. **Validation**: Evaluated clustering quality using WSS and silhouette analysis
6. **Visualization**: Created heatmaps, PCA plots, and cluster profile visualizations
7. **Enrichment Analysis**: Performed pathway enrichment analysis to identify biological functions

## 9.2 Key Findings

From our analysis of the example dataset:

- Identified `r length(unique(hc_result$clusters))` distinct gene clusters using hierarchical clustering
- Cluster sizes ranged from `r min(table(hc_result$clusters))` to `r max(table(hc_result$clusters))` genes
- Found `r nrow(enrichment_summary)` significantly enriched pathways across clusters
- PCA revealed clear separation between experimental conditions

## 9.3 Next Steps

Based on these results, researchers can:

- Investigate specific genes within interesting clusters
- Validate clustering results with additional datasets
- Perform detailed functional analysis of enriched pathways
- Design follow-up experiments based on cluster-specific patterns

## 9.4 Session Information

```{r session_info}
sessionInfo()
```

**Note**: This vignette uses simulated data for demonstration purposes. When working with real gene expression data, ensure proper quality control, consider batch effects, and validate biological interpretability of results.